---
title: "Solutions to tidying exercises"
author: "Ina Bornkessel-Schlesewsky"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Exercises from R4DS

### 12.2 Exercise 2: computing the rate from table2 and table4

Compute the rate for table2 first:

```{r rate_table2}
cases_per_year <- table2 %>%
  filter(type =="cases")

population_per_year <- table2 %>%
  filter(type =="population")

rate <- cases_per_year$count / population_per_year$count * 10000
country <- rep(unique(table2$country), each=2)
year <- rep(unique(table2$year), 3)
type <- rep("rate",6)

interim_table <- tibble(country,year,type,rate) %>% 
  rename(count = rate)

newtable2 <- table2 %>% 
  bind_rows(interim_table) %>% 
  arrange(country,year,type)

newtable2
```

Now compute the rate for table4a (which includes the cases) and table4b (which includes the population):

```{r rate_table4}
pop1999 <- table4b$`1999`
pop2000 <- table4b$`2000`

newtable4 <- table4a %>% 
  mutate(`1999` = `1999` / pop1999 * 10000,
         `2000` = `2000` / pop2000 * 10000)
```

### 12.2 Exercise 3: plot cases over time from table2

```{r plot_from_table2}
ggplot(filter(table2,type =="cases"), aes(year, count)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country)) +
  ylab("cases")

```

Note that the following syntax also works:

```{r plot_from_table2_alt}
table2 %>% 
  filter(type == "cases") %>% 
  ggplot(aes(year, count)) + 
    geom_line(aes(group = country), colour = "grey50") + 
    geom_point(aes(colour = country)) +
    ylab("cases")

```

### 12.3 Exercise 1

Why are pivot_longer() and pivot_wider() not perfectly symmetrical?
Carefully consider the following example (Hint: look at the variable types and think about column names.):

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```

*This is, in part, a sorting issue. Also, year is a character column when reconverted due to the values having been column names after conversion to wider format.*

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return") %>% 
  arrange(year,half)
```

pivot_longer() has a names_ptypes argument, e.g.  names_ptypes = list(year = double()). What does it do?

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return", names_ptypes = list(year = double()))
```

### 12.3 Exercise 2

Why does this code fail?

```{r}
table4a %>% 
  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")

```

### 12.3 Exercise 3

What would happen if you widen this table? Why? How could you add a new column to uniquely identify each value?

```{r}
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)

people %>% 
  pivot_wider(names_from = names, values_from = values)
```

