---
title: "Importing data"
subtitle: "Data Science for Cognitive (Neuro)Science"
author: "Ina Bornkessel-Schlesewsky"
date: "Jan 2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(emo)
```

---

# NYC squirrels revisited!

<br>

* Typically, the data we work with aren't available as R packages
* Rather, we need to read them in from a file
* csv is a popular format
  * comma separated values
  * plain text
  * can be exported from all popular spreadsheet applications (e.g. Excel)
  * (+ is accessible without any proprietary software, allows for version control etc. -- more on this later ...)
* Let's try importing the squirrels data from a csv file (i.e. in the form in which it can be downloaded from the *NYC Open Data* website)

---

# Importing a csv file

* there are various ways to do this in R
* we will use the `read_csv()` function from the `readr` package (part of the `tidyverse`)

```{r import_squirrels}

squirrels <- read_csv("data/2018_Central_Park_Squirrel_Census_-_Squirrel_Data.csv")

```

---

# Inspect the data

```{r inspect_squirrels}
head(squirrels, n = 2)
```

* interactively in RStudio, you can also use `view(squirrels)` for a spreadsheet-type visualisation

---

# Variable types

* R guesses the variable type based on the first 1000 observations
* Common types include *dbl* (double, a number); *chr* (character, often also referred to as "strings"), *lgl* (logical, TRUE/FALSE)
* We may want to convert some character variables to factors (i.e. categorical variables with known set of possible values, "levels") at a later point
* Columns that only contain numbers will be classed as *dbl*, as soon as any entry contains even a single non-numeric character, the column will be classed as *chr* (cf. "Unique Squirrel ID")
  * this is important to know if, for any reason, your numerical variables contain characters, e.g. to code missing values; `NA`, which is R's internal code for missing values, is an exception
  * note also that `read_csv` has correctly filled in `NAs` in the squirrels data where there were empty cells
* If you have a large dataset where the values change in some way after the first 1000 observations, this might produce an error later on due to a misclassification of your variables

---

# Challenges I: variable names

* Some variables have spaces in their names
* This is cumbersome, as we would always have to enclose the variable name in single backward quotes
* Try out what happens when you omit the quotation marks or use a different type of quotation marks

```{r}
squirrels %>%
  filter(`Primary Fur Color` == "Gray")

```

---

# Challenges I: variable names

* We can fix this by renaming the variable using the `rename` function
* Note how I've changed the spelling of "colour" at the same time

```{r}
squirrels <- squirrels %>%
  rename(primary_fur_colour = `Primary Fur Color`)

squirrels %>%
  filter(primary_fur_colour == "Gray")
```

---

# Challenges II: values

* Why are the values for date so strange (e.g. *1.01e7*)?
* This is clearly not a date!

---

# Challenges II: values

.pull-left[
* Examining the original csv-file (or using `view`) reveals that dates are coded simply as numbers using the American MMDDYYYY ordering convention, e.g. "10142018" for October 14th 2018
* Conveniently, the *lubridate* package provides a range of functions for creating date/time objects
]

.pull-right[
```{r}
library(lubridate)

mdy(10142018)

squirrels <- squirrels %>%
  mutate(Date = mdy(Date))

squirrels$Date
```
]

---

# A bit more "wrangling"

* To recreate some of the analyses that we looked at earlier, we'll rename the relevant variables and restrict the data frame to only include those variables

```{r}
squirrels <- squirrels %>%
  rename(hectare = Hectare,
         shift = Shift,
         date = Date,
         age = Age) %>%
  select(hectare, shift, date, age, primary_fur_colour)

squirrels

```

---



---
class: inverse, mline, center, middle

# R programming basics

The following closely mirrors R4DS chapter 2 (section 4 in the online version of the book)
---

